Geospatial - Assignment 1

### 2025-01-27 hw1/submission_notebook.Rmd
Paper 3, markdown opticals - Refined paper 3 - Text and optical changes to .rmd

@@

@@

```{r}
library(osmdata)
library(sf)
library(spData)
library(tidyverse)
library(geodata)
library(readxl)
library(ggplot2)
library(raster)
library(terra)
library(lwgeom)
library(geobr)
```

## [Paper 1](https://www.sciencedirect.com/science/article/pii/S0304387818300166#appsec1)

### Data loading

The authors of the paper provide a link to the data, however, the link is unfortunately outdated. Yet, the data is available in a newer version on the website (authors used 2013 data, we use 2024 data).

@@

### Data wrangling

@@

### Plotting

@@

## [Paper 2](https://www.sciencedirect.com/science/article/pii/S0166046220303100)

### Data loading

Unfortunately, again the data was not directly available. Therefore, we leverage some r packages to get the data for highways, powergrids (osmdata) and population data from 2020 (geodata). The ESSR survey data was directly available via the World Bank data repositories as a CSV (2011/12) data used as in the paper. We load all of these in the following:

@@

### Data wrangling

We calculate the population density for each district and clip the roads to the Ethiopia boundary.

@@

## [Paper 3](https://papers.ssrn.com/sol3/papers.cfm?abstract_id=3776990)

### Data loading

@@

We retrieved population data from the aforementioned IBGE website and used geographical data available in the R package geobr. While the authors reproduced the same plot for three different years — 1950, 1980, and 2010 — due to a lack of data, we were only able to reproduce the plot for the most recent year. Specifically, on the IBGE website, we found data for 2009. However, we assume no significant variations in population occurred within a single year, allowing us to proceed with our analysis.

Additionally, we explored the R package censobr, which is expected to contain Brazil's census data for earlier years. Unfortunately, despite using the function data_dictionary and experimenting with various arguments (e.g., "population," "households," etc.), we were unable to retrieve population data. Instead, the package provided other types of data that were not relevant to our analysis.

```{r}
# Read the .xls file with the Brasilian population for 2009
df <- read_excel("data/paper3/UF_Municipio.xls", skip = 4, col_names = TRUE)

# Get the geographical data from the package geobr
state <- read_state(year = 2010)
meso <- read_meso_region(year = 2010)
region <- read_region(year = 2010)
```

### Data wrangling

From the package geobr, data at the meso-regions level are available, but they have a different level of granularity compared to the population data. This discrepancy can be easily verified by comparing the number of observations in the two datasets. Due to this inconsistency between the geographical data and the population data, we could not exactly replicate the authors' plots, which show the Brazilian population shares for meso-regions. Instead, we had to perform the analysis at the state level.

First, we grouped the population data by state to ensure that both datasets referred to the same political boundaries. Next, we joined the two datasets based on the state code. After this, we created a column indicating the population share, multiplying the ratio by 100 to make the results more consistent with the paper. Finally, we filtered the data for the North and Central-West regions to highlight the borders in the plot.

```{r}
# Convert ESTIMADA (=population) into a numerical variable and # handle missing values by setting them to 0
df <- df %>%
  mutate(ESTIMADA = as.numeric(ESTIMADA)) %>%
  mutate(ESTIMADA = replace_na(ESTIMADA, 0))

# Group by the state code
df_grouped <- df %>%
  group_by(COD...2) %>%
  summarise(
    Total_Estimada = sum(ESTIMADA, na.rm = TRUE),
    Count = n()
  )

# Remove the last row of df_merged - in df the last row was the # total and it would have been considered another meso region
df_grouped <- df_grouped %>%
  slice(-n())

# Merge the geographical data with the population data
df_merged <- state %>%
  left_join(df_grouped, by = c("code_state" = "COD...2"))

# Create the column pop_share
df_merged <- df_merged %>%
  mutate(pop_share = Total_Estimada / sum(Total_Estimada) * 100)

@@

### Plotting

```{r}
# Final plot
ggplot(data = df_merged) +
  ggtitle("(c) 2010") +
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_sf(aes(fill = pop_share), color = "white") +
  scale_fill_gradientn(
    colors = c("lightblue", "blue", "darkblue"),
    name = "Pop. Share",
    limits = c(0, 25)
  ) +
  geom_sf(data = region, aes(color = factor(name_region)), fill = NA, linewidth = 1) +
  scale_color_manual(values = c("black", "red")) +
  guides(color = "none")
```

@@

Again, as outlined above, we only replicate the 2010 map, due to data non-availability in the prior periods.

## Paper 4

### Data loading

### Data wrangling

### Plotting

@@

## Paper 5

### Data loading

### Data wrangling

### Plotting

>>>>

<hr>

```{css, echo=FALSE}
h1.title {
  text-align: center;
}

h4.author, h4.date {
  text-align: center;
}

.author {
  text-align: center;
}

.date {
  text-align: center;
}
```

@@

<div style="text-align: center;">
  <img src="Barcelona_School_of_Economics_logo.svg" alt="BSE logo">
</div>

<hr>

@@

```{r}
# Main tidyverse (includes ggplot2)
library(tidyverse)

# Spatial packages
library(sf)
library(terra) # Newer replacement for raster
library(osmdata) # OpenStreetMap data
library(spData) # Spatial data
library(geobr) # Brazilian boundaries
library(censobr) # Brazilian population data
library(lwgeom) # Needed for st_make_valid

# Data import
library(geodata) # For elevation data
library(readxl) # For Excel files
```

<hr>

# [Paper 1](https://www.sciencedirect.com/science/article/pii/S0304387818300166#appsec1)

## Data loading

The authors of the paper provide a link to the data, however, the link is unfortunately outdated. Yet, the data is available in a newer version on the [website](https://www.dws.gov.za/DSO/publications.aspx) (authors used 2013 data, we use 2024 data) but had to be converted to geojson format before usage.

@@

## Data wrangling

@@

## Plotting

@@

<hr>

# [Paper 2](https://www.sciencedirect.com/science/article/pii/S0166046220303100)

## Data loading

Unfortunately, again the data was not directly available. Therefore, we leverage some r packages to get the data for highways, powergrids (osmdata) and population data from 2020 (geodata). The [ERSS survey](https://microdata.worldbank.org/index.php/catalog/2053/data-dictionary) data was directly available via the World Bank data repositories as a CSV (2011/12) data used as in the paper. We load all of these in the following:

@@

## Data wrangling

We calculate the population density for each district and clip the roads to the Ethiopia boundary. Furthermore, t looks like the authors used slightly different road and powergrid networks. However, overall the grids look broadly similar when utilizing OpenStreetMap data.

@@

<hr>

# [Paper 3](https://papers.ssrn.com/sol3/papers.cfm?abstract_id=3776990)

## Data loading

@@

We retrieved population data from the aforementioned IBGE website and used geographical data available in the R package geobr. While the authors reproduced the same plot for three different years — 1950, 1980, and 2010 — due to a lack of data, we were only able to reproduce the plot for the most recent year, i.e., 2010. Specifically, on the IBGE website, we found data for 2009, and the censobr package provides the 2010 data. It also provides census data for 1980 and 1960, but in 1980, it seems the population weights are missing. Therefore, we decided to focus on the 2010 data.

For geographical boundaries of Brazil for administrative areas, we use the geobr package.

```{r}
# Read in the Brasilian population for 2010 using the censobr package
# This also contains 1980, and 1960 data, but 1980 without weights, and 1960 only partially
pop_2010 <- censobr::read_population(
  year = 2010,
  columns = c("code_state", "code_region", "code_muni", "V0010"),
  showProgress = FALSE,
  as_data_frame = TRUE,
  # cache = FALSE
) %>% # Group by municipality code
  group_by(code_muni) %>%
  summarise(
    total_population = sum(V0010, rm.na = TRUE)
  )

# Get the geographical data from the package geobr and validate if necessary
meso <- read_meso_region(year = 2010) %>%
  st_make_valid()
region <- read_region(year = 2010)
muni <- read_municipality(year = 2010) %>%
  st_make_valid()
```

## Data wrangling

First, we grouped the population data by state to ensure that both datasets referred to the same political boundaries. Next, we joined the two datasets based on the meso code. We had to take a rather coarse-granular approach by using the `st_centroid()` function to match meso-regions with municipalities. this leads us to loosing out roughly on 50'000 inhabitants. After this, we created a column indicating the population share, multiplying the ratio by 100 to make the results more consistent with the paper. Finally, we filtered the data for the North and Central-West regions to highlight the borders in the plot. Needing `sf_remove_holes()` to avoid inner borders, we loose a bit of granularity in the outer borders of the broader regions.

```{r}
# First, join the population data with municipal boundaries
pop_spatial <- muni %>%
  left_join(pop_2010, by = "code_muni")

# Use st_centroid to ensure each municipality is counted only once
pop_meso <- pop_spatial %>%
  st_centroid() %>% # Convert polygons to points
  st_join(meso) %>% # Join with meso regions
  group_by(code_meso) %>%
  summarise(
    total_population = sum(total_population, na.rm = TRUE)
  ) %>%
  st_drop_geometry() # Remove geometry before joining back

# Now join back with meso geometries
pop_meso <- meso %>%
  left_join(pop_meso, by = "code_meso")

# Create pop_share
df_merged <- pop_meso %>%
  mutate(pop_share = total_population / sum(total_population) * 100)

@@

## Plotting

```{r}
ggplot(data = df_merged) +
  ggtitle("(c) 2010") +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5), # Center title while keeping minimal theme
    panel.grid = element_line(color = "gray90"), # Make grid lines lighter
    axis.text = element_blank(),
    axis.ticks = element_blank()
  ) +
  geom_sf(aes(fill = pop_share), color = "white", size = 0.1) +
  scale_fill_stepsn(
    name = "Pop. Share",
    breaks = c(0, 0.216, 0.382, 0.557, 1.02, 11.12),
    labels = scales::number_format(accuracy = 0.001),
    colors = c("#eff3ff", "#bdd7e7", "#6baed6", "#3182bd", "#08519c"), # Made lower values darker
    values = scales::rescale(c(0, 0.216, 0.382, 0.557, 1.02, 11.12))
  ) +
  geom_sf(data = region, aes(color = factor(name_region)), fill = NA, linewidth = 1) +
  scale_color_manual(values = c("black", "red")) +
  guides(color = "none") +
  # Add titles
  labs(
    caption = "Source: Based on Pellegrina and Sotelo (2021)"
  )
```

@@

Again, as outlined above, we only replicate the 2010 map, due to data issues in the prior periods.

<hr>

# Paper 4

## Data loading

## Data wrangling

## Plotting

@@

<hr>

# Paper 5

## Data loading

## Data wrangling

## Plotting
